Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-01 WSSSPE1_report/WSSSPE1_summary_report.tex

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
\emph{FASTMath}.  Most of the above efforts are focused on one scientific
and/or applied mathematics domain. However, sometimes such projects grow beyond
the initial vision of developers. One such example is \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. An instance of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
becoming a general purpose utility system usable for solving variety of
scientific problems.
%

%Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?
In order to investigate the much discussed issue of career paths for developers
of scientific software, we conducted a small survey of the use-cases discussed
above with respect to the current status of its core developers. We looked at
the workshop submissions and the original cited papers for the respective
software. Where available, we looked at the webpages and source code of the
earliest versions to find the original authors. We excluded relatively new
projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
that in a majority of scientific software more than one original developers are
still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
Whereas, in the case of utility software projects, it appears that at least one
original developer is still around often in a part-time role. In other cases,
it appears that the original developers have went on to assume a larger role,
again with the same project.

From this survey, it appears that the career paths for utility software
developers are broader compared to the developers of niche software. However,
as projects as a whole, the small-scale, specific software have a lower risk
and cost in terms of sustaining as compared to utility software. Often the
sustainability of such projects is achieved by the fact that the core
developer or team heavily utilizes the software for their science, eg.
\emph{R/qtl, PETSc}. Furthermore, the development of scientific software requires more
scientific skills compared with those of utility software. This diminishes the
need of specialist programmers and practices in many cases where the bulk of
development is carried out by a domain scientist.

>>>>

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
\emph{FASTMath}.  Most of the above efforts are focused on one scientific
and/or applied mathematics domain. However, sometimes such projects grow beyond
the initial vision of developers. One such example is \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. An instance of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
becoming a general purpose utility system usable for solving variety of
scientific problems.
% 
In order to investigate the much discussed issue of career paths for developers
of scientific software, we conducted a small survey of the use-cases discussed
above with respect to the current status of its core developers. We looked at
the workshop submissions and the original cited papers for the respective
software. Where available, we looked at the webpages and source code of the
earliest versions to find the original authors. We excluded relatively new
projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
that in a majority of scientific software more than one original developers are
still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
Whereas, in the case of utility software projects, it appears that at least one
original developer is still around often in a part-time role. In other cases,
it appears that the original developers have went on to assume a larger role,
again with the same project.

In conclusion, we find that there are distinctions in the characteristics and
the kinds of challenges faced between generic utility and specific scientific
domain software projects. In terms of career paths of original developers, it
appears that the career paths for utility software developers are broader
compared to the developers of scientific software.  However, it appears that in
the case of scientific software, core competencies of original developers tend
to be domain science who happen to be the coders of their software. As projects
as a whole, the small-scale, specific software have a lower risk and cost in
terms of sustaining as compared to utility software.  Often the sustainability
of such projects is achieved by the fact that the core developer or team
heavily utilizes the software for their science, eg.  \emph{R/qtl, PETSc}.
Furthermore, the development of scientific software requires more scientific
skills compared with those of utility software. This diminishes the need of
specialist programmers and practices in many cases where the bulk of
development is carried out by a domain scientist. With these inferences, we
believe that separate guidelines and/or principles for sustainability be
defined and practiced for each category of scientific software and they be
given separate treatment.
